Writing Trash Bin

### 2024-03-13 paper.qmd
rendered for meeting
finshed rewrite of mistakes made section

## Similar work
::: aside
I have a large backlog of uncertainty literature review summaries in my Mendely that need to be added here. There are a lot of them.
:::

@@

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

@@

## Mistakes when we misunderstand uncertainty
We feel the need to make this point clear because a *large* amount of uncertainty research seems to be confused by the purpose of uncertainty visualisation, what it is related to, and why we visualise it. Often, the authors are even more confused about what they are doing than the participants. There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations. Instead, we have a handful of highlights presented below.  
- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.
- @Hofman2020 asked questions about discusses the difference in showing inferential or outcome (prediction) intervals and then found that participants shown the inferential distribution performed poorly when asked questions about predictions. They found that showing the relevant distribution to the question reduced the error. This result should be astoundingly obvious if it was well understood that uncertainty is not some latent variable and is only relevant for specific ground truth statistics. This papers *percieved* importants highlights an underlying belief that uncertainty is linked to a specific variable or data set, not the variable we are drawing inference on.  
- @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant. @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the core ideas of these papers, the fact that they were published indicates that the general population was not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values. This indicates a general failure to see bias and variance both as sources of uncertainty.  
- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.  
- @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertaninty information". These papers fail to realise the exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis. Therefore, it is performed to *suggest* a direction for future inference work or communication, which *do* contain uncertainty and does not contain uncertainty within itself. This confusion would not occur if it was understood that uncertainty does not exist without inference.  
- @Boukhelifa2012 tried to quantify the strength of the intuitive connection between "sketchiness" and uncertainty. Since they had seen literature that explained uncertainty was task specific, they tested the connection for multiple "tasks", what they failled to understand, is that the "task" dependency which is likely just a proxy of the evaluation experiment problem discussed in the previous section. This misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty. I imagine it is quite hard to be "uncertain" about the existence of a train line. Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.

@@

The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship.

This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty. The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty. This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of uncerstanding and quantifying them are ad-hoc at best. @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy. None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication. Realistically, this problem is well known, and put simply by Freeman Dyson in his fameous Birds and Frogs speech:

> Rigorous theorems are the best way to give a subject intellectual depth and precision. Until you can prove rigorous theorems, you do not fully understand the meaning of your concepts. - Freeman Dyson (Birds and Frogs speech)

This limitation becomes apparent when you look at the limited scope of what the uncertainty visualisation literatrue has *actually* focused on visualising. Almost every uncertainty visualisation paper contains a broad and encapsulating definition of uncertainty in the introduction, followed by a visualisation of a normally distributed probablity density function.

> Edit notes: Need some way to show uncertainty visualisation papers provide sweeping definitions and then visualise a simple PDF everytime.

In order to visualise something, it needs to be quantifiable. In order to quanitfy something, it needs to be mathematically defined. The broad classes of what is consdiered "uncertainty" is not currently quantifiable in any way that is not ad-hoc. The only apsects of uncertaitny that are currently quantifiable are confidence intervals, prediction intervals, and related terms. Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis. Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.  This means that the current literature of uncertainty visualisation papers are doing one of two things:  
1) Evaluating multiple visual representations of a single mathematical expression of uncertainty (e.g. a PDF, confidence interval or error). These papers are purposefully vague, and would be clearer in their intention if the title was aligned with the actual mathematical object they are visualising. For example, a paper that is titled "evaluation of uncertainty visualisations" that only considers PDF visualisation should realistically be titled "evaluation of PDF visualisations".  
2) Evaluating multiple visual representations of *multiple* different mathematical expression of uncertainty. Since these experiments visualise different mathematical objects, some of which are closer to the ground truth statistic, the results of these papers can be easily predicted with an understanding of how to get the ground truth statistic from a plot and the perceptual tasks required to do so.

It is clear that the vague definition is causing across the board misunderstandings of uncertainty, confusing experiment methodologies and motivations, and a literature that is difficult to navigate. This is also causing a large body of experiments that have easily predictable results that can be easily explained by the existing visualisation. This problem will be the focus of the following section.

- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming. While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.

# What is an Uncertainty Visualisation?
The lack of a consistent definition of uncertainty means that every statement that builds upon "uncertainty" will be ill-formed. This includes statements about uncertainty visualisation. This appears in the literature as a mountain of seemingly conflicting statements. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [] (*Cite Gap __*).

Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem. For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue. @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy. What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may beleive this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.

@@

## Uncertainty Visualisation is not defined

@@

@@

# Noise and Signal
## Uncertainty versus Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

@@

## Decision Making

Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

@@

## Other attemps to capture uncertainty
There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].  

@@

## Uncertainty is not hypothesis testing
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

@@

## Suggestions for measuring uncertainty
### Types of Uncertainty Visualisation Papers

>>>>

## Similar work

@@

The idea that inference is related to uncertainty pops up in most discussions of uncertainty, however, due to the now *implicit* understanding we are performing inferential statistics, it is often brought up as a *task* or *goal* dependence. This is mentioned both by authors at a specific stage of an analysis, and by authors looking at an entire field or at all stages of the analysis pipeline. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. These examples show that authors that discuss uncertainty believe it is there is an important relationship between uncertainty and what we decide to keep in our analysis and throw away, i.e. the task or goal. The task or goal is, in essence, the *statistic you are drawing inference on*. Were this not true, we would have no reason to differentiate between a prediction interval and a sampling distribution because they could both be considered "uncertainty about the mean", however they are different because their inferential goals are different.

@@

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty *when we perform inference*. These things are not "uncertainty" in of themselves, but rather contribute to the distance between the final estimate and the statistic we want to perform inference on. The field of uncertainty communication is in desperate need of a unifying mathematical definition that firmly identifies exactly how each of these things contribute to uncertainty. This may be a considerable task, but as of right now, the list of things that are "uncertainty" continues to grow, and the best methods of understanding and quantifying them are ad-hoc at best. @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, information theory tries to quantify uncertainty using the idea of entropy. None of these existing methods are thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication. The need for a mathematical definition of this field is put simply by Freeman Dyson in his fameous Birds and Frogs speech:

> "Rigorous theorems are the best way to give a subject intellectual depth and precision. Until you can prove rigorous theorems, you do not fully understand the meaning of your concepts." @Dyson2010

In order to visualise something, it needs to be quantifiable. In order to quanitfy something, it needs to be mathematically defined. The broad classes of what is consdiered "uncertainty" is not currently quantifiable in any way that is not ad-hoc. The only apsects of uncertaitny that are currently quantifiable are confidence intervals, prediction intervals, and related terms. Even then, these quantifications often only capture the uncertainty in our model or its assumptions, it ignores bias often introduced in earlier stages of the analysis. Broader and more complicated concepts such as the effects of assumptions, imputed missing variables, and model choices remain difficult to meaningfully quantify beyond ad-hoc methods.

## Mistakes when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence). When uncertainty visualisaiton authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself. In these cases, the authors are more confused about what they are doing than the participants.

There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations. Instead, we have a handful of highlights.

@Hofman2020 performed an experiment where some participants were shown a sampling distribution of the mean, and others were shown a prediction interval around the mean, for the effectiveness of a particular treatment. The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval. A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectivness of the treatment*. It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017]. The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* statistics. This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.

@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. The authors of this study were aware that there was some kind of task dependence with respect to displaying uncertainty, so they tested the connection for multiple "tasks". The authors did not seem to understand the "task" dependence is likely a dependence on the inference for a particular statistic, so the "task" dependence was interpreted as *context* dependence. @fig-sketchy depicts the six scenarios that participants were shown and asked to interpret what they believe the squiggly line indicates. The authors misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty. It seems a tad ridiculous for the creators of a rail network map to be "uncertain" about the existence of a train line, however we are unable to guess at what the authors of the paper believed the uncertainty would be. The most likely scenario is that the authors did not consider the specific estimate that the uncertainty was representing, which is what led to visualisations that do not make conceptual sense. Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.

@@

Many authors do not understand bias *is* uncertainty (something that has been proven mathematically by @Meng2014) and whether we view uncertainty as bias or variance depends on the resolution of our problem. @Padilla2021 found that high uncertainty in the model estimates (calculated uncertainty) and low forecaster confidence (which is typically an expression of suspected bias in the model) both caused participants to have decreased confidence in their results and suggested modelers express both if they are relevant. @Kale2019 discussed the importance of communicating decisions made in the data analysis pipeline and being aware of the alternatives. While there is nothing wrong with the results of these papers, the fact that they were published shows some level of surprise by these results. This work indicates that there was a significant number of authors in the field were not already aware that choices introduced early in the data analysis pipeline create bias and therefore uncertainty in our final values. The surprise that both bias and variance contribute to final uncertainty shows that authors do not understand both will contribute to the *distance* of your estimate to the value you are drawing inference on (AKA error).

The misunderstanding that descriptive statistics also extends to visualisation. The exploration step of an analysis, which includes descriptive statistics, exploratory data visualisaiton, and unsupervised machine learning techniques, is performed without a prior hypothesis, however misunderstandings of this fact appears frequently in the literature. @Griethe2006 commented that "if visualization is used as a means to explore a data volume or to communicate its contents the uncertainty has to be included". @Potter2010 aimed to create a summary plot that "concisely presented data with uncertainty information" to create an exploratory visualisation tool that visualised uncertainty.

This confusion (or indifference) also creates many imprecisely named papers. Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out (*Cite Gap 9*). While this is a minor by product of the definition issue discussed earlier, it does result in a literature that is needlessly difficult to navigate.

The reality is that a lot of uncertainty visualisation authors do not seem to have an intuitive understanding of uncertainties connection to inference, when inference is being performed, and how to design experiments that capture this relationship. This misunderstanding is not due to laziness or the fault of the authors, but rather is likely caused by the absence of a strict definition of uncertainty.The imprecise and confusing set of existing definitions are creating a field in which the authors themselves do not know what they are testing.

# What is an Uncertainty Visualisation?
The lack of a consistent definition of uncertainty means that every statement that builds upon "uncertainty" will be ill-formed. This includes statements about uncertainty visualisation. This appears in the literature as a mountain of seemingly conflicting statements. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [@Ibrekk1987] (*Cite Gap __*). 
Cite Gap: Confusion about what "uncertainty visualisation" is
- @Wickham2011 suggests their product plot framework, which includes histograms, should have a way to measure uncertainty, but does not consider that a histogram is *already* a depiction of mass and would already be considered an uncertainty visualisation were our statistic of interest the population mean. To suggest that a histogram needs uncertainty bounds is similar to suggesting a density or box plot need uncertainty bounds.

Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem. For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue. @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy. What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may believe this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.

@@

## Mathematically incomparable objects

@@

Cite gap __: represent different maths objects in visualisations
- @wu2023rational recognised it is difficult to understand when differences in performance stem from information asymmetry.

@@

# Noise and Signal

## Types of Uncertainty Visualisation Papers

## Uncertainty versus Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

## Types of Uncertainty Visualisation Papers

@@

## Decision Making
Decision making questions that typically involve uncertainty discuss the concept of "risk-aversion" as though it is a mistake. The irony of an uncertainty visualisation pper expressing the importance of uncertainty visualisation, but setting up a ground truth or optimum strategy in which participants *should* ignore the uncertainty information is not lose on us. This perspective does not make sense because it implies participants should make decisions entirely based on the average. A possible soltion for this is to provide a utility framework for a particular experiment [@Hullman2016], however it is unclear how easily participants could adopt a utility framework that is different to their own. @Cheong2016 introduced a very simple optimum strategy into their experiment on bushfire uncertainty visualisations, where participants should choose to evacuate a house as soon as its likelihood of catching fire is 50% or higher. Interestingly, despite this very obvious and simple tactic to maximize payout from the experiment, it seems like many participants did not adopt it, instead acting much more warily as though they were considering whether or not they would *actually* evacuate in the event of a fire. These papers also very rarely consider what the optimum decision would be in a given scenario, and a difficulty with anticipating the best strategy is conflated with the failings of the uncertainty visualisation. Ultimately, we agree with @Hullman2016 who suggested that what determines an approripate ground truth is largely a philosophical exercise.

@@

## Other attemps to capture uncertainty
There are other studies that try to include uncertainty information without directly asking for it but do not go down the route of asking for a decision or levels of trust. These papers often try to ask a question that should utilise both the uncertainty and signal in the response. This is an interesting approach, but these papers are often still looking for a "signal" which is specified in the form of a "ground truth" that is used to evaluate the plot. Unfortunately this usually comes in the form of slightly cryptic questions that create a large amount of noise on the interpretation side [@Hullman2016].

Cite gap __: Weird questions to capture uncertainty

@@

## Uncertainty and hypothesis testing
It is unclear how participants are supposed to incorperate uncertainty information into their responses when uncertainty visualisation authors themselves are unsure. Other authors interpret this difficulty as a combination of the uncertainty and task complexity [@Kinkeldey2014], and not as a general confusion. All these examples sever to illustrate that *we dont know how to measure uncertainty* and if we don't know how to measure it, *we cannot evaluate the quality of a visualisation*. It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis, but this idea also has a miriad of issues that come with it.  There is already a series of issues with the reject/do not reject concepts in hypothesis testing, so it is already questionable to set the results of a hypothesis test as the ground truth in an uncertainty visualisation experiment, however these may be additional problems that go beyond the binary nature of the issue. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. This increased sensisitivity could be the result of another aspect of uncertainty that is ignored. Statstical tests are typically built upon a series of assumptions, and it is difficult to identify *which* assumption failure caused the data to fail the statistical tests. Lineup plots allow us to see *how* our data is different to the assumed distribution, and better understand *why* our data may have failled a test ^[Did Patricks paper compare statistical tests to lineup protocols in the event an earlier assumption of the tests, such as, independent observation? I think he did check some. I wonder if you can quantify the information difference you get from a rejected statistical test vs a rejected line plot]. These experiments utilised the lineup protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypyvs shows the conceptual difference between the lineup protocol and uncertainty visualisation. A lineup protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significiantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). ^[Does this flip have mathematical implications in how the plots should be assessed?] This implies a visible signal in an uncertainty plot should indicate some divergence from the null.

@@

## Suggestions for measuring uncertainty

### 2024-03-13 writingdumpster.qmd
finshed rewrite of mistakes made section

@@

# Main section outtakes
- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"

>>>>

Cite Gap 9: Many papers will boast a title that claims to be about uncertainty visualisaiton, but simply depicts different visual representations of a PDF, however it takes reading the methodology to find this out

@@

# Main section outtakes
- The process of working out whether or not a visualisation shows a piece of information, and if it does, what perceptual task it is mapped to and what assumptions or heuristics need to be used, is *incredibly* time consuming. While this method does predict the results for these experiments with remarkable accuracy, to do it for a large number of plots, especially when the ground truth statistics and visualisation are incredibly complicated is just not feasible for a reviewer to do.
- This is likely because The idea that uncertainty can only be defined in the presence of a motivating question is well grounded in most areas of statistics. The entire process of data analysis, from deciding what should be observed as data through to communicating that data in a plot is governed by human decision and the goal of an analysis.
- We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"
